\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-27 tex/paper.tex
save

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likehood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variationial upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In this note, we develop a likelihood-free inference algorithm for the point
estimation from observed data of the parameters of a forward non-differentiable
generative model. We propose to adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an abitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

We consider a family of parameterized densities $p(\mathbf{x}|\theta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p(\mathbf{x}|\theta) = \int p(\mathbf{x}|\mathbf{z},\theta) p(\mathbf{z}|\theta) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in {\cal Z}$ is a latent variable providing an external
source of randomness. In particular, $\mathbf{z}$ is not necessarily assumed to
be a fixed-size vector (e.g., it can be a sequence of variable length) and its
distribution $p(\mathbf{z}|\theta)$ may itself depend on $\theta$ in some intricated way.

@@

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q(\theta|\psi) \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \theta_m \sim q(\theta|\psi), \mathbf{z}_m \sim p(\mathbf{z}|\theta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{ \theta_m \sim q(\theta|\psi),  \mathbf{z}_m \sim p(\mathbf{z}|\theta_m) \}_{m=1}^M$.}  \Comment{Update $q_\psi$}
            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
            % \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$}
            \State{$\psi \leftarrow \text{Adam}(\nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p(\mathbf{x}|\psi) \equiv \theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q(\theta|\psi)$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
More formally, the learned model $p(\mathbf{x}|\psi)$ therefore corresponds to the marginal distribution
$\int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta$ of the generated data.
\glnote{Connection with variational bayes?}

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q(\theta|\psi)i$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
proposal distributions with larger entropy are not penalized, which may result
in learning a smeared variation of the original simulator.
\glnote{Add note on minimum entropy, otherwise it degenerates back to a non-differentiable function.}

@@

\subsection{Gaussian mixture model}

@@

\subsection{Physics example}

@@

As reviewed in \cite{2016arXiv161003483M}, likelihood-free inference is
intimately tied to a class of algorithms that can be framed as density
estimation-by-comparison. In most cases, these inference algorithms are
formulated as an iterative two-step process where the model distribution is
first compared to the true data distribution and then updated to make it more
comparable to the latter. Closest to our work are procedures that rely on a
classifier to estimate the discrepancy between the true and the generated data
distributions. For example, \citep{gutmann2012noise} uses non linear logistic
regression for fitting unnormalized differentiable statistical models, while
\citep{goodfellow2014generative} exploits an adversarial neural network for
learning a differentiable implicit generative model. In the likelihood-free
setup, \citep{cranmer2015approximating,2016arXiv161110242D} estimate likelihood
ratios through supervised classification, which can in turn be used for
parameter inference in combination with a gradient-free optimization algorithm.
Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
a summary statistics for approximate Bayesian computation.
\glnote{Explain what's new about the proposed work.}

>>>>

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likelihood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In this note, we develop a likelihood-free inference algorithm for the point
estimation from observed data of the parameters of a forward non-differentiable
generative model. We propose to adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

We consider a family of parameterized densities $p(\mathbf{x}|\theta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p(\mathbf{x}|\theta) = \int p(\mathbf{x}|\mathbf{z},\theta) p(\mathbf{z}|\theta) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in {\cal Z}$ is a latent variable providing an external
source of randomness. In particular, $\mathbf{z}$ is not necessarily assumed to
be a fixed-size vector (e.g., it can be a sequence of variable length) and its
distribution $p(\mathbf{z}|\theta)$ may itself depend on $\theta$ in some intricate way.

@@

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q(\theta|\psi) \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \theta_m \sim q(\theta|\psi), \mathbf{z}_m \sim p(\mathbf{z}|\theta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{ \theta_m \sim q(\theta|\psi),  \mathbf{z}_m \sim p(\mathbf{z}|\theta_m) \}_{m=1}^M$.}  \Comment{Update $q(\theta|\psi)$}
            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
            % \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$}
            \State{$\psi \leftarrow \text{Adam}(\nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p(\mathbf{x}|\psi) \equiv \theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q(\theta|\psi)$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
More formally, the learned model $p(\mathbf{x}|\psi)$ therefore corresponds to the marginal distribution
$\int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta$ of the generated data.

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q(\theta|\psi)i$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
proposal distributions with larger entropy are not penalized, which may result
in learning a smeared variation of the original simulator.
\glnote{Add note on minimum entropy, otherwise it degenerates back to a non-differentiable function.}

@@

\subsection{Toy problems}

As illustrative experiments, we evaluate in this section adversarial variational
optimization on well-known distributions for continuous and discrete data.

We first consider inference for a univariate Gaussian with unknown mean and
unknown variance. The observed data is sampled from a univariate Gaussian with
mean $\mu = 1$ and variance $\sigma^2 = 0.5^2$. Algorithm~\ref{alg:avo} is run
for 300 epochs with mini-batches of size $M=64$ and the following configuration.
For the critic, we use a 3-layer MLP with 10 hidden nodes per layer and ReLU
activations. At each epoch, Adam is run for $n_\text{critic}=100$ iterations
with a step size $\alpha=0.01$, decay rates $\beta_1=\beta_2=0.5$ and its inner
first and second moment vectors reset at each epoch in order to avoid building
momentum in staled directions.  For modeling $\mu$ and $\sigma^2$, we use
univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
\sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
parameters $\psi$ are updated by making one Adam step, with $\alpha=0.01$ and
$\beta_1=\beta_2=0.1$. Penalty coefficients are set to $\lambda=0.025$ and
$\gamma=5$.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
    \caption{\glnote{todo}}\label{fig:gaussian}
\end{figure}

The upper plot of Figure~\ref{fig:gaussian} illustrates the resulting model
after the optimization procedure. As expected from adversarial training, the
model density $p(\mathbf{x}|\psi)$ closely matches the true density
$p_r(\mathbf{x})$, as shown by the aligned histograms in blue and orange. Under
the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we also observe
that the proposal distributions have correctly concentrated their mass around
$\mu=1$ and $\log \sigma=-0.69$. The bottom plot of Figure~\ref{fig:gaussian}
shows an empirical estimate of $-U_d$ with respect to the epoch. We observe
that it quickly falls towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$
and that the critic cannot distinguish between true and model data.

Similarly, ... \glnote{discrete case with a Poisson distribution.}

@@

@@

As reviewed in \cite{2016arXiv161003483M}, likelihood-free inference is
intimately tied to a class of algorithms that can be framed as density
estimation-by-comparison. In most cases, these inference algorithms are
formulated as an iterative two-step process where the model distribution is
first compared to the true data distribution and then updated to make it more
comparable to the latter.

Closest to our work are procedures that rely on a classifier to estimate the
discrepancy between the true and the model distributions. For example,
\citep{gutmann2012noise} uses non linear logistic regression for fitting
unnormalized differentiable statistical models, while
\citep{goodfellow2014generative} exploits an adversarial neural network for
learning a differentiable implicit generative model. In the likelihood-free
setup, \citep{cranmer2015approximating,2016arXiv161110242D} estimate likelihood
ratios through supervised classification, which can in turn be used for
parameter inference in combination with a gradient-free optimization algorithm.
Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
a summary statistics for approximate Bayesian computation.

In this context, the proposed method can be considered as a direct adaptation of
generative adversarial networks~\citep{goodfellow2014generative} to
non-differentiable simulators. It also constitutes an approximate gradient
descent alternative to bayesian optimization based on density
ratios~\citep{cranmer2015approximating} or to classifier
ABC~\citep{gutmann2017likelihood}.
